|

### 2024-08-26 reproducibility/manuscript/manuscript.qmd
starting to update methods, part 2

First, we introduce the generative model that describes the various unobserved times:
\begin{align}
  % unit lognormal t_c
  t_c &\sim \text{LogNormal}(0, 1) \\
  % gene-specific t_0
  t^{(0)}_{0,g} &\sim \text{LogNormal}(0, 1) \\
  % switching time
  \Delta \textrm{switching}_g &\sim \text{LogNormal}(0, 1) \\
  % gene-specific t_1
  t^{(1)}_{0,g} &= t^{(0)}_{0,g} + \Delta \textrm{switching}_g \\
  %cell-gene-specific activation state
  k_{cg} &\sim \text{Bernoulli}(\textrm{logits}=t_c - t^{(1)}_{0,g}) \\
  % cell-gene-specific latent time
  \tau_{cg} &= \text{softplus}(t_c - t^{(k_{cg})}_{0,g}).
\end{align}
Here, $\tau_{cg}$ represents the displacement of time per
cell and gene with 
\begin{align}
 textrm{softplus}(t) :=  \log( 1 + e^t).
\end{align}
Recall that $t_c$ is the shared time per cell, $t^{(k_{cg})}_{0,g}$ is the
gene-specific switching time. Each cell and gene combination has its
transcriptional state $k_{cg} \in \{ 0, 1 \}$, where $0$ indicates the
activation state and $1$ indicates the expression state. Each gene has two
switching times for representing activation and repression: $t^{(0)}_{0,g}$ is
the first switching time corresponding to when the gene expression starts to be
activated, $t^{(1)}_{0,g}$  is the second switching time corresponding to when
the gene expression starts to be repressed.

Next we introduce the priors for the splicing parameters (where the activation rate $\alpha$ depends on the activation state $k_{cg}$ from above):
\begin{align}
  \alpha^{(0)}_g &\sim \text{LogNormal}(0, 1) \\
  \beta_g &\sim \text{LogNormal}(0, 1) \\
  \gamma_g &\sim \text{LogNormal}(0, 1) \\
  \alpha_{cg} = \begin{cases}
    \alpha^{(0)}_g & \text{if } k_{cg} = 0 \\
    0 & \text{if } k_{cg} = 1
  \end{cases}
\end{align}
\textbf{Note that $\alpha^{(1)}$ is shared for all the genes, while ${\alpha^{(0)}}_g$ is learned independently for
each gene. MATT: this was in the old text, but I think $\alpha^{(1)}$ is no longer used based on conversations with Alvin?}

@@

Finally, we describe the ODE solution at time $\tau_{cg}$, and the observation model at those times that give rise to the observed counts:
\begin{align}
  \hat{u}_{cg}, \hat{s}_{cg} &= \textrm{ODESolve}\Big( u^{(0)}_{cg}, s^{(0)}_{cg}, \alpha_{cg}, \beta_g, \gamma_g; \ T_0=0, T_1=\tau_{cg} \Big) \\
  u_{cg} &\sim \text{Poisson}(\hat{u}_{cg}) \\
  s_{cg} &\sim \text{Poisson}(\hat{s}_{cg})
\end{align}

The analytic solution of the differential equations to predict
spliced and unspliced gene expression given their parameters is derived by the
authors of scVelo and a theoretical RNA velocity study
[@Bergen2020-pj;@Li2021-qa] and given in
Eqs. \ref{eq-solution-u}-\ref{eq-solution-s2}.
\begin{align}
u\left(\tau^{\left(k_{c g}\right)}\right) 
  &= u_0^{\left(k_{c g}\right)}{ }_g e^{-\beta_g \tau^{\left(k_{c g}\right)}} 
  \nonumber \\
&\hskip -24pt + \frac{\alpha^{\left(k_{c g}\right)}}
  {\beta_g}\left(1-e^{-\beta_g \tau^{\left(k_{c g}\right)}}\right) 
  \label{eq-solution-u}\\
s\left(\tau^{\left(k_{c g}\right)}\right) 
  &= s_0^{\left(k_{c g}\right)} e^{-\gamma_g \tau^{\left(k_{c g}\right)}} 
  \nonumber \\
  &\hskip -24pt + \frac{\alpha^{\left(k_{c g}\right)}}{\gamma_g}
    \left(1-e^{-\gamma_g \tau^{\left(k_{c g}\right)}}\right) 
    \nonumber\\
  &\hskip -24pt + \frac{\alpha^{\left(k_{c g}\right)}-\beta_g u_0^{\left(k_{c g}\right)}}
    {\gamma_g-\beta_g}\left(e^{-\gamma_g \tau^{\left(k_{c g}\right)}}
    -e^{-\beta_g \tau^{\left(k_{c g}\right)}}\right), 
    \nonumber \\
  &\qquad \beta \neq \gamma \label{eq-solution-s} \\
s\left(\tau^{\left(k_{c g}\right)}\right) 
  &= {s_0^{\left(k_{c g}\right)}}_g e^{-\beta_g \tau^{\left(k_{c g}\right)}} \nonumber \\
  &\hskip -24pt +\frac{\alpha^{\left(k_{c g}\right)}}{\beta_g}
  \left(1-e^{-\beta_g \tau^{(k c g)}}\right) 
  \nonumber \\
  &\hskip -24pt -\left(\alpha^{\left(k_{c g}\right)}
  -\beta_g u_0^{\left(k_{c g}\right)}{ }_g\right) \tau^{\left(k_{c g}\right)} 
  e^{-\beta_g \tau^{\left(k_{c g}\right)}}, \nonumber \\
  &\qquad \beta = \gamma \label{eq-solution-s2}
\end{align}

@@

We use these solutions to formulate an end-to-end probabilistic generative model 
that relates prior distributions on kinetic parameters to a distribution on pairs 
of observed unspliced and spliced read count matrices

\begin{align}
\alpha^{(0)}{ }_g &\sim \operatorname{LogNormal}(0,1), \\
\beta_g &\sim \operatorname{LogNormal}(0,1), \\
\gamma_g &\sim \operatorname{LogNormal}(0,1), \\
&\hskip -18pt \Delta \text { switching }_g \sim \operatorname{LogNormal}(0,1), \\
t_0^{\left(k_{c g}\right)} &= \left\{
  \begin{array}{l}
    t_0^{(0)}{ }_g \sim \operatorname{Normal}(0,1), k_{c g}=0 \\
    t_0^{(1)}{ }_g=t_0^{(0)}{ }_g+\Delta \text { switching }_g, \\
    \quad k_{c g}=1
  \end{array}\right. \\
t_c &\sim \operatorname{LogNormal}(0,1), \\
k_{c g} &\sim \text{Bernoulli} \left( \text{logits}= t_c-t_0^{(1)} \right), \\
\tau^{\left(k_{c g}\right)} 
  &= \operatorname{softplus}\left(t_c-t_0^{\left(k_{c g}\right)}{ }_g\right), \\
u_{c g} 
 &= \text { Measurement }_u \left( u\left(\tau^{\left(k_{c g}\right)}\right) ; 
                                   u_{c g}^{obs}\right), \\
s_{c g} 
  &= \text { Measurement }_s \left( s\left(\tau^{(k_{c g})}\right) ; 
                                    s_{c g}^{obs}\right).
\end{align}
$u\left(\tau^{\left(k_{c g}\right)}\right)$ and $s\left(\tau^{(k_{c g})}\right)$ are
are called the denoised gene expression calculated from the velocity analytic
solution input with the kinetics random variables. $u_{cg}$ and $s_{cg}$ are the spliced
and unspliced read count sampled from the Poisson models. $u_{cg}^{obs}$ and $s_{cg}^{obs}$ are
the observed spliced and unspliced read count tables. The generative process

$\text{Measurement}(\cdot)$ for observed unspliced read counts given denoised unspliced
gene transcript expression level $u\left(\tau^{(k_{cg})}\right)$ 
(and identical for observed spliced read counts) models the expected number of observed 
reads for a given gene in a given cell as the number of transcripts times the ratio of 
the cell's total reads to total transcripts
\begin{align}
u_c^{\hat{obs}} &= \sum_g u_{c g}^{obs}, \\
\hat{u}_c &= \sum_g u\left( \tau^{(k_{c g})}\right), \\
\eta_c^{(u)} &\sim \operatorname{Normal}\left(
    u_c^{\hat{obs}_c}, 
    \operatorname{std} \left(u_c^{\hat{obs}}\right)
  \right), \\
\mu_{c g}^{(u)} &= \log \left(u\left(\tau^k{ }_{c g}\right)\right)
  +\log \left(\eta_c^{(u)}\right)-\log \left(\hat{u}_c\right), \\
u_{c g}^{obs} &\sim 
  \operatorname{Poisson}\left(\lambda=\exp \left(\mu_{c g}^{(u)}\right)\right).
\end{align}

For the first Pyro-Velocity model (Model 1), we constrain the shared time to be strictly
larger than $t_{0}^{(0)}$ by introducing auxiliary random variables 
$$
\text{t\_constraint}_{cg} 
  \sim \text{Bernoulli} \left( \text{logits} = t_c - {t_{0}^{(0)}}_g \right),
$$
and setting their values to $1$, and we set the initial condition per gene to be
\begin{align}
\left( {u_{0}^{(k_{cg})}}_g , {s_{0}^{(k_{cg})}}_g \right) &= \left\{
  \begin{array}{l}
    (0,0), k_{c g}=0 \\
    \bigg( {u \left( \Delta \text { switching }_g \right)}_g,\\
           \quad {s \left( \Delta \text { switching }_g \right)}_g \bigg), \\
           \quad k_{c g}=1
  \end{array}\right.
\end{align}
For the extended Pyro&thinsp;-Velocity model (Model 2), we remove the shared
time constraint $\text{t\_constraint}_{cg}$, thus allowing a time lag per gene
that might be caused by delayed gene activation and set the initial condition
per gene as random variables that are strictly positive $\left({u_{0}^{(0)}}_g,
{s_{0}^{(0)}}_g\right)$, which allow genes having a basal expression level before gene
activation. Then, we compute the gene expression at the second switching time as
\begin{align}
({u_{0}^{(1)}}_g, {s_{0}^{(1)}}_g) &= 
  \bigg( {u \left( \Delta \text { switching }_g \right)}_g, \nonumber \\
& \qquad {s \left( \Delta \text { switching }_g \right)}_g \bigg),
\end{align}
which shares the same initial condition $\left({u_{0}^{(0)}}_g, {s_{0}^{(0)}}_g\right)$ where
\begin{align}
{u_{0}^{(0)}}_g &\sim \operatorname{LogNormal}(0,1),\\
{s_{0}^{(0)}}_g &\sim \operatorname{LogNormal}(0,1).
\end{align}

>>>>

First, we introduce the generative model that describes the various unobserved times:
\begin{align}
  % unit lognormal t_c
  t_c &\sim \text{LogNormal}(0, 1) \\
  % gene-specific t_0
  t^{(0)}_{0,g} &\sim \text{LogNormal}(0, 1) \\
  % switching time
  \Delta \textrm{switching}_g &\sim \text{LogNormal}(0, 1) \\
  % gene-specific t_1
  t^{(1)}_{0,g} &= t^{(0)}_{0,g} + \Delta \textrm{switching}_g \\
  %cell-gene-specific activation state
  k_{cg} &\sim \text{Bernoulli}(\textrm{logits}=t_c - t^{(1)}_{0,g}) \\
  % cell-gene-specific latent time
  \tau_{cg} &= \text{softplus}(t_c - t^{(k_{cg})}_{0,g}).
\end{align}
Here, $\tau_{cg}$ represents the displacement of time per
cell and gene with 
\begin{align}
 \text{softplus}(t) :=  \log( 1 + e^t).
\end{align}
Recall that $t_c$ is the shared time per cell, $t^{(k_{cg})}_{0,g}$ is the
gene-specific switching time. Each cell and gene combination has its
transcriptional state $k_{cg} \in \{ 0, 1 \}$, where $0$ indicates the
activation state and $1$ indicates the expression state. Each gene has two
switching times for representing activation and repression: $t^{(0)}_{0,g}$ is
the first switching time corresponding to when the gene expression starts to be
activated, $t^{(1)}_{0,g}$  is the second switching time corresponding to when
the gene expression starts to be repressed.

Next we introduce the priors for the splicing parameters (where the activation rate $\alpha$ depends on the activation state $k_{cg}$ from above):
\begin{align}
  \alpha^{(0)}_g &\sim \text{LogNormal}(0, 1) \\
  \beta_g &\sim \text{LogNormal}(0, 1) \\
  \gamma_g &\sim \text{LogNormal}(0, 1) \\
  \alpha_{cg} &= \begin{cases}
    \alpha^{(0)}_g & \text{if } k_{cg} = 0 \\
    0 & \text{if } k_{cg} = 1
  \end{cases}
\end{align}
\textbf{Note that $\alpha^{(1)}$ is shared for all the genes, while ${\alpha^{(0)}}_g$ is learned independently for
each gene. MATT: this was in the old text, but I think $\alpha^{(1)}$ is no longer used based on conversations with Alvin?}

@@

Finally, we describe the ODE solution at time $\tau_{cg}$, and the observation model at those times that give rise to the observed counts:
\begin{align}
  \hat{u}_{cg}, \hat{s}_{cg} &= \text{ODESolve}\Big( u^{(0)}_{cg}, s^{(0)}_{cg}, \alpha_{cg}, \beta_g, \gamma_g; \ T_0=0, T_1=\tau_{cg} \Big) \\
  \mu^{(u)}_c &= \sum_{g=1}^G {u}^{\text{(obs)}}_{cg}, \quad \mu^{(s)}_c = \sum_{g=1}^G {s}^{\text{(obs)}}_{cg} \\
  \sigma^{(u)}_c &= \sqrt{\frac{1}{G} \sum_{g=1}^G \left( u_{cg}^{\text{(obs)}} - \mu^{(u)}_c \right)^2} \\
  \sigma^{(s)}_c &= \sqrt{\frac{1}{G} \sum_{g=1}^G \left( s_{cg}^{\text{(obs)}} - \mu^{(s)}_c \right)^2} \\
  \eta^{(u)}_c &\sim \text{Normal}\Big(\mu^{(u)}_c, \ \sigma^{(u)}_c\Big) \\
  \eta^{(s)}_c &\sim \text{Normal}\Big(\mu^{(s)}_c, \ \sigma^{(s)}_c\Big) \\
  \hat{\mu}^{(u)}_c &= \sum_{g=1}^G \hat{u}_{cg}, \quad \hat{\mu}^{(s)}_c = \sum_{g=1}^G \hat{s}_{cg} \\
  \lambda^{(u)}_{cg} &= \log(\hat{u}_{cg}) + \log(\eta^{(u)}_{c}) - \log(\hat{\mu}^{(u)}_c) \\
  \lambda^{(s)}_{cg} &= \log(\hat{s}_{cg}) + \log(\eta^{(s)}_{c}) - \log(\hat{\mu}^{(s)}_c) \\
  \hat{u}^{\text{(obs)}}_{cg} &\sim \text{Poisson}\Big(\exp (\lambda^{(u)}_{cg})\Big) \\
  \hat{s}^{\text{(obs)}}_{cg} &\sim \text{Poisson}\Big(\exp (\lambda^{(s)}_{cg})\Big)
\end{align}
Here, we use ${u}^{\text{(obs)}}_{cg}, {s}^{\text{(obs)}}_{cg}$ to denote the observed unspliced and spliced counts
for cell $c$ and gene $g$. We use $\hat{u}^{\text{(obs)}}_{cg}, \hat{s}^{\text{(obs)}}_{cg}$ to
denote our generative model's prediction of these unspliced and spliced expression levels.
The generative process for modeling these observed read counts given denoised gene transcript expression level $\hat{u}_cg, \hat{s}_cg$ considers the expected number of observed reads for a given gene in a given cell as the number of transcripts times the ratio of the cell's total reads to total transcripts.
\textbf{Improve descriptions of how noise is modeled in the observation model.}

\textbf{Need to update the analytic solutions, but first need to confirm the above is correct. Also, I recommend pushing all of the below analytic solutions to the appendix.}
The analytic solution of the differential equations to predict
spliced and unspliced gene expression given their parameters is derived by the
authors of scVelo and a theoretical RNA velocity study
[@Bergen2020-pj;@Li2021-qa] and given in
Eqs. \ref{eq-solution-u}-\ref{eq-solution-s2}.
\begin{align}
u\left(\tau^{\left(k_{c g}\right)}\right) 
  &= u_0^{\left(k_{c g}\right)}{ }_g e^{-\beta_g \tau^{\left(k_{c g}\right)}} 
  \nonumber \\
&\hskip -24pt + \frac{\alpha^{\left(k_{c g}\right)}}
  {\beta_g}\left(1-e^{-\beta_g \tau^{\left(k_{c g}\right)}}\right) 
  \label{eq-solution-u}\\
s\left(\tau^{\left(k_{c g}\right)}\right) 
  &= s_0^{\left(k_{c g}\right)} e^{-\gamma_g \tau^{\left(k_{c g}\right)}} 
  \nonumber \\
  &\hskip -24pt + \frac{\alpha^{\left(k_{c g}\right)}}{\gamma_g}
    \left(1-e^{-\gamma_g \tau^{\left(k_{c g}\right)}}\right) 
    \nonumber\\
  &\hskip -24pt + \frac{\alpha^{\left(k_{c g}\right)}-\beta_g u_0^{\left(k_{c g}\right)}}
    {\gamma_g-\beta_g}\left(e^{-\gamma_g \tau^{\left(k_{c g}\right)}}
    -e^{-\beta_g \tau^{\left(k_{c g}\right)}}\right), 
    \nonumber \\
  &\qquad \beta \neq \gamma \label{eq-solution-s} \\
s\left(\tau^{\left(k_{c g}\right)}\right) 
  &= {s_0^{\left(k_{c g}\right)}}_g e^{-\beta_g \tau^{\left(k_{c g}\right)}} \nonumber \\
  &\hskip -24pt +\frac{\alpha^{\left(k_{c g}\right)}}{\beta_g}
  \left(1-e^{-\beta_g \tau^{(k c g)}}\right) 
  \nonumber \\
  &\hskip -24pt -\left(\alpha^{\left(k_{c g}\right)}
  -\beta_g u_0^{\left(k_{c g}\right)}{ }_g\right) \tau^{\left(k_{c g}\right)} 
  e^{-\beta_g \tau^{\left(k_{c g}\right)}}, \nonumber \\
  &\qquad \beta = \gamma \label{eq-solution-s2}
\end{align}

@@
